---
title: "analysis"
author: "Zikang Ge"
date: "2024-12-12"
output: html_document
---
```{r}
library(dplyr)
library(tidyverse)
library(ggplot2)
library(reshape2)
library(wordcloud)
library(RColorBrewer)
```

```{r}
url <- "https://github.com/TheGooseGuy/RShiny-Spotify-Dashboard/raw/main/spotify-2023.csv"
dataset <- read_csv(url)
```

```{r}
str(dataset)
summary(dataset)

```
```{r}
# Fill missing numeric values with the median
dataset <- dataset %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), median(., na.rm = TRUE), .)))

# Check again for missing values
colSums(is.na(dataset))
```
```{r}
# Standardize column names for consistency
colnames(dataset) <- tolower(gsub("[[:punct:]]", "_", colnames(dataset)))

# Basic statistics for numeric columns
numeric_summary <- dataset %>%
  summarise(across(where(is.numeric), list(
    mean = ~mean(. , na.rm = TRUE),
    median = ~median(. , na.rm = TRUE),
    sd = ~sd(. , na.rm = TRUE)
  )))

numeric_summary
```
```{r}
# Count of songs by release year
ggplot(dataset, aes(x = as.factor(released_year))) +
  geom_bar(fill = "#1DB954", color = "black") +
  labs(title = "Number of Songs Released by Year",
       x = "Release Year", y = "Count") +
  theme_minimal()+
    

# Count of songs by release month
ggplot(dataset, aes(x = as.factor(released_month))) +
  geom_bar(fill = "#FF5733", color = "black") +
  labs(title = "Number of Songs Released by Month",
       x = "Release Month", y = "Count") +
  theme_minimal()

```

```{r}
# Visualize release trends
# Create a heatmap of release trends
heatmap_data <- dataset %>%
  count(released_month, released_day)

ggplot(heatmap_data, aes(x = as.factor(released_month), y = as.factor(released_day), fill = n)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "white", high = "#1DB954") +
  labs(title = "Heatmap of Song Releases", x = "Release Month", y = "Release Day") +
  theme_minimal()

```

```{r}
# Distribution of danceability
ggplot(dataset, aes(x = danceability__)) +
  geom_histogram(binwidth = 5, fill = "#FF5733", color = "black") +
  labs(title = "Distribution of Danceability",
       x = "Danceability (%)", y = "Count") +
  theme_minimal()

# Distribution of energy
ggplot(dataset, aes(x = energy__)) +
  geom_histogram(binwidth = 5, fill = "#1DB954", color = "black") +
  labs(title = "Distribution of Energy",
       x = "Energy (%)", y = "Count") +
  theme_minimal()

# Distribution of valence
ggplot(dataset, aes(x = valence__)) +
  geom_histogram(binwidth = 5, fill = "#FFC300", color = "black") +
  labs(title = "Distribution of Valence",
       x = "Valence (%)", y = "Count") +
  theme_minimal()

```

# K means clustering

```{r}
# Select numerical features for clustering
features <- dataset %>%
  select(bpm, danceability__, energy__, valence__, acousticness__, liveness__)

# Normalize the features
features_scaled <- as.data.frame(scale(features))

# Check the first few rows of the scaled features
head(features_scaled)
```

```{r}
# Determine optimal number of clusters
# Calculate total within-cluster sum of squares for different cluster counts
wss <- sapply(1:10, function(k) {
  kmeans(features_scaled, centers = k, nstart = 10)$tot.withinss
})

# Plot the Elbow Method
plot(1:10, wss, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of Clusters", ylab = "Total Within-Cluster Sum of Squares",
     main = "Elbow Method for Optimal Clusters")

```

```{r}
# K means clustering
# Apply k-means clustering
set.seed(42) # For reproducibility
kmeans_result <- kmeans(features_scaled, centers = 4, nstart = 10)

# Add the cluster assignment to the dataset
dataset$cluster <- as.factor(kmeans_result$cluster)

```



```{r}
# Perform PCA on scaled features
pca <- prcomp(features_scaled)

# Create a data frame with PCA results and clusters
pca_data <- data.frame(PC1 = pca$x[, 1], PC2 = pca$x[, 2], Cluster = dataset$cluster)

# Plot clusters
ggplot(pca_data, aes(x = PC1, y = PC2, color = Cluster)) +
  geom_point(alpha = 0.7, size = 2) +
  labs(title = "Clusters of Songs (PCA Visualization)", x = "Principal Component 1", y = "Principal Component 2") +
  theme_minimal()

```


```{r}
# box plot of feature distribution by clusters
# Boxplot for danceability by cluster
ggplot(dataset, aes(x = cluster, y = danceability__)) +
  geom_boxplot(fill = "#1DB954", alpha = 0.7) +
  labs(title = "Danceability by Cluster", x = "Cluster", y = "Danceability (%)") +
  theme_minimal()

# Boxplot for energy by cluster
ggplot(dataset, aes(x = cluster, y = energy__)) +
  geom_boxplot(fill = "#FF5733", alpha = 0.7) +
  labs(title = "Energy by Cluster", x = "Cluster", y = "Energy (%)") +
  theme_minimal()
```
```{r}
# Calculate average feature values per cluster
cluster_summary <- dataset %>%
  group_by(cluster) %>%
  summarise(
    avg_bpm = mean(bpm),
    avg_danceability = mean(danceability__),
    avg_energy = mean(energy__),
    avg_valence = mean(valence__),
    avg_acousticness = mean(acousticness__),
    avg_liveness = mean(liveness__)
  )

print(cluster_summary)


```


Based on the average feature values of each cluster, we can assign descriptive labels to better represent the "vibes" of the songs in each cluster:

Cluster Descriptions
Cluster 1:

Description: High tempo, moderate danceability, and energy, but low valence (happiness). Likely represents intense or moody tracks, possibly from genres like trap, electronic, or dark pop.
Suggested Label: Intense & Dark
Cluster 2:

Description: Moderate tempo with high danceability, energy, and valence. These tracks are vibrant, upbeat, and likely to be feel-good songs from genres like pop, dance, or happy electronic.
Suggested Label: Upbeat & Happy
Cluster 3:

Description: Moderate tempo, good energy, and valence with higher liveness. Songs in this cluster could be live recordings or performances, and likely fall into acoustic, indie, or live pop categories.
Suggested Label: Energetic & Live
Cluster 4:

Description: Low energy and danceability but very high acousticness. Likely represents slower, acoustic, or calmer genres such as folk, ambient, or ballads.






